Twenty years of dynamic occupancy models: a review of applications and look towards the future

### 2024-09-27 Manuscript.qmd
Completed final version

---
title: "Twenty years of dynamic occupancy models: a review of applications and a look towards the future"
date: today

@@

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a key objective of much ecological research. Nonetheless, reliably estimating occupancy can be a challenge, particularly when common issues such as imperfect detection and shifting distributions are at play. The dynamic, multi-season occupancy model (DOM) offers a framework for occupancy estimation which explicitly accounts for observation error, while also incorporating additional mechanistic process by explicitly estimating colonisation and extinction rates. In the two decades since its introduction, the DOM has become a popular tool for describing patterns in occupancy, exploring the environmental drivers of species occurence, and, more recently, for predicting species occupancy to new locations and to future conditions.

  In this review we examine how ecologists have applied the DOM in its first two decades of use. Based on a sample of articles fitting DOMs to field ecological data, we explore the systems in which these models have been used, how data were collected for them, and how authors have built their models to achieve their objectives. Our findings indicate that the DOM is an particularly flexible tool readily applied to a range of diverse taxa from around the globe, capable of being used to estimate occupancy at study sites ranging from local to continental scales over time periods from months to decades. The DOM framework is also amenable to extension, further broadening their utility to address a wide range of ecological questions.

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
---

@@

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing is applicable in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy difficult to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

@@

In this review we examine how dynamic occupancy models have been used by ecologists in the two decades since their inception. Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied the DOM to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models. Based on these results we highlight the DOM's flexibility as a tool for understanding species occurrence, examine approaches to the model building process, and outline key priorities for future research involving this model class.

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied (z = 1) or unoccupied state (z = 0). In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε (@eq-occupancy).

$$
\begin{align*}
z_1 \sim& Bernoulli(\psi_1)\\\\
z_t \sim& Bernoulli(\psi)\\\\
&\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
\end{align*}
$$ {#eq-occupancy}

In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0). Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur (@eq-detection).

$$
y_{t,j} \sim Bernouilli(z_t\rho)
$$ {#eq-detection}

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process submodel describes changes in occupancy over time, where occupancy in the first season is given by the probability of initial occupancy and occupancy in ensuing seasons is governed by colonisation and extinction probabilities. The observational process submodel describes detectability, where the probability of detecting a species given that a site is occupied is given by the detection probability. At sites which are truly unoccupied, it is assumed that the species is never detected (i.e., no false positives).](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017a; @guillera-arroita2017b]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this enables flexibility in data inputs. In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform, @eq-occupancy, and @eq-detection), though some alternative model forms may directly estimate occupancy ψ as a parameter.

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

@@

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites. A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@kéry2021].

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.

Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles that did not meet inclusion criteria were replaced from within their own strata.

As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.

@@

A total of 92 articles were included for this review. Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).

![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata. The proportion of articles included from each strata were as follows: 12.5% from 2005-2008; 24.4% from 2008-2011; 42.6% from 2012-2015; 35.1% from 2016-2019; and 57.1% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how the noted variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranged from under one month to forty years, with a median of 6 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

The DOM has been used to study a variety of different species, though the majority of studies have focussed on vertebrate taxa (@fig-StudyDetails B). The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications. @mores2020 and @padilla-torres2013 have used the DOM to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats. The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon [@fisher2014], Great Plains stream fishes [@falke2012], and whales [@pendleton2022].

While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the degree to which models are tailored to the taxa is likely to be limited for the sake of practicality.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 55 articles fit models to only a single taxon, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa. C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as emerging methods like camera traps and acoustic monitors. 13 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

The presence/absence data required for DOMs has come from a breadth of different sources and detection methods (@fig-StudyDetails, panel C). In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps. Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location. For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 use aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.

Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling. In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@warrier2020; @pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the "certain" detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. The interpretation of what 'occupancy' means and the factors which drive it may differ depending on the scale of what is considered a site, and how that relates to the ecology of a species. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. The DOM may be readily applied to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data. For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days. While this provides interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.

@@

    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography, including hydrology</i>"),

>>>>

---
title: "Twenty years of dynamic occupancy models: a review of applications and look towards the future"
date: today

@@

abstract: | 
 Understanding the patterns by which species occur across landscapes and throughout time is a long-standing objective of ecological research which has inspired the development of numerous quantitative modelling approaches for the task. However, estimating occupancy can be a challenge, particularly where issues like imperfect detection and shifting distributions exist. Dynamic occupancy models (DOMs) offer a framework for occupancy estimation which explicitly accounts for observation error while incorporating additional mechanistic process by estimating colonisation and local extinction. In light of increasing interest in more process-explicit models for understanding species occurrence, here we examine how DOMs have been applied to field ecological data in the two decades since their introduction. Following a general introduction to the model, we present the results of a systematic review exploring the research contexts where DOMs have been applied and interrogating how authors have built their models, with particular emphasis on how covariates are incorporated to describe variation in occupancy dynamics. Our findings indicate that DOMs are a flexible tool readily applied to diverse study systems and data types, with their usage expanding in recent years as more studies apply them to make spatial and temporal prediction of species occupancy. DOMs are also amenable to extension, further broadening their utility. Despite their flexibility, model complexity in DOMs tends to be low; most studies consider relatively few covariates, most of which are represented as simple linear relationships. Approaches to covariate selection also vary considerably, and there remains little research on how these choices may influence model performance. Furthermore, only a fraction of articles report evaluating models and little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
---

@@

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013]. Robust knowledge of the occupancy patterns of a species can also help us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and quantitatively describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly applied to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy difficult to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

@@

In this review we examine how dynamic occupancy models have been used by ecologists in the two decades since their inception. Following an introduction to the model's form and assumptions, we present the results of a systematic review exploring how researchers have applied DOMs to ecological data, with emphasis on how they collected their data, selected covariates, and evaluated their models. Based on these results we highlight the DOM's flexibility as a tool for understanding species occurrence, examine approaches to the model building process, and outline key priorities for future research involving this model class.

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites can exist in either an occupied (z = 1) or unoccupied state (z = 0). In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~, i.e., $z_1 \sim Bernouilli(\psi_1)$. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε, such that $z_t \sim Bernoulli(\psi)$ and $\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma$. The observation component of the DOM explicitly accounts for imperfect detection. At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0), according to the formula $y_{t,j} \sim Bernouilli(z_t\rho)$. Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur. In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform). However, some alternative model forms may directly estimate occupancy ψ as a parameter.

![The form of the dynamic occupancy model as described by @mackenzie2003. The ecological process sub-model describes changes in occupancy over time, while the observational process sub-model describes detectability.](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017a; @guillera-arroita2017b]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this enables flexibility in data inputs.

![The sampling design of the standard dynamic occupancy model. During seasons (also called primary occasions) sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations (also called surveys or secondary occasions) are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. Note that it is not necessary for each site or season to have the same number of observations, and that missing data can be accommodated.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

@@

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and it can be difficult to distinguish local extinction from temporary emigration [@valente2017]. Model extensions to relax the closure assumption have been developed, including approaches using staggered arrival and departure periods between sites [@kendall2013]. A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@steenweg2018].

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters (ψ~1~, γ, ε, and ρ) to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species can depend not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates [@barry2006]. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.

Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. These queries resulted in 1469 articles: 897 were retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles that did not meet inclusion criteria were replaced from within their own strata.

As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) used non-simulated, field-collected, presence-absence data; ii) relied on data from multiple sites which could exist in at least two states, including an occupied and unoccupied state; iii) had multiple seasons, between which sites could change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contained at least one parameter describing the detection process.

@@

A total of 92 articles were included for this review. Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).

![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata. The proportion of articles included from each strata were as follows: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}

The articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of attributes of these studies is presented in @fig-StudyDetails, and throughout this section we provide notable examples of DOMs to highlight how the noted variation appears in practice. Study systems included in our sample are globally distributed: while a majority of articles use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems (@fig-StudyDetails A). A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale (@fig-StudyDetails B) . The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The time elapsed between the first and last survey ranged from under one month to over forty years, with a median duration of 8 years (@fig-StudyDetails F). More meaningful, however, is the period of the primary occasion, as this represents the temporal scale at which changes in occupancy are assumed to occur. Most applications describe primary occasions as occurring annually, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week and capture much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. One example of this is @kleiven2020's study of rodent and mustelid interactions using camera trap data, which used primary occasions less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France which used data from two primary occasions: one in 2003 and one in 2012.

DOMs have been used to study a variety of species, though the majority of studies have focused on vertebrate taxa (@fig-StudyDetails C). The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions, including DOMs used to model decadel changes in lichen occupancy and the spread of chronic wasting disease [@belinchón2017; @cook2022]. The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications [@bailey2014]. Multiple authors have also used DOMs to model mosquito dynamics, a strongly applied use-case with important human health implications [@mores2020; @padilla-torres2013]. The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon, Great Plains stream fishes, and whales [@fisher2014; @falke2012; @pendleton2022].

While most studies in our sample fit a model to a single species, 44% fit models to multiple species either as independent models (34%) or explicitly multi-species implementations (12% — some articles attempted multiple approaches; @fig-StudyDetails D). The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the degree to which models are tailored to the taxa is likely to be limited for the sake of practicality.

![A) Locations of study areas where data was collected for reviewed DOMs. B) Spatial extent of study areas, defined as the area of inference, within which all surveyed points were contained. C) Number of articles which fit models to each category of taxa. Taxa were considered 'threatened' if they are listed on the IUCN Red List, or if authors indicate that they are otherwise threatened. D) Explicitly multi-species models include both hierarchical, jointly estimated models as well as more interactive models. Some studies fit both independent and multi-species models, such that these values do not sum to our sample size. E) Survey methods used to collected presence/absence data. Note that some articles employed multiple detection methods, and that some methods (e.g., citizen bird counts) may fall into multiple categories. F) Quantity of sites where surveys were conducted and duration of studies. Yellow bars indicate median values for site quantity (100) and study duration (8.2 years). Study duration is defined as the time elapsed between the first and last survey.](Figures/ReviewBox.png){#fig-StudyDetails}

Studies collected the presence/absence data required for DOMs from a breadth of different sources and detection methods (@fig-StudyDetails E). In our sample, 79% of studies used human observational data, 11% used live-trapping methods, and 14% used detections from camera traps. 10% of articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods to determine the presence of a species at a location. For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy; in another unique implementation, @marescot2020 fit a unique 'multi-species' model treating poachers as a taxon and using ranger reports to create detection histories.

Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling. In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which predate the DOM, as in a study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data; some integrate multiple sources of data to maximise sample size. In one example, @warrier2020 modelled tiger occupancy in India using three sources of detections, including camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the "certain" detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. The interpretation of what 'occupancy' means and the factors which drive it may differ depending on the scale of what is considered a site, and how that relates to the ecology of a species. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. The DOM may be readily applied to these data types, and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data [@balantic2019]. This is already apparent in recent camera trap studies which collapsed their detections into seasons of just a few days, far shorter than is realistic with conventional survey methods [@kleiven2020; @mölle2022]. While these studies provide interesting insights of occupancy at extremely fine temporal scale, further research could help provide general guidance on determining appropriate season and survey durations with respect to research questions and the closure assumption.

@@

    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
    "HYDR" ~ html("<b>Hydrology</b>\n<i>Aspects of hydrology, such as distance to water</i>"),
